---
title: "Designing and running a randomized trial with PhysioTrial"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing and running a randomized trial with PhysioTrial}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

PhysioTrial provides reproducible trial-design infrastructure: randomization with
allocation concealment, CONSORT/SPIRIT operations, explicit analysis
populations, power and sample-size helpers, and deterministic CDISC-shaped
exports. This vignette walks one small trial end to end on synthetic data.

These exports are auditable and structurally checked; they are **not** a claim of
formal CDISC, third-party validator, or regulatory acceptance.

```{r setup}
library(PhysioTrial)
```

## 1. Define the trial

```{r trial}
trial <- Trial(
  "rehab-01",
  arms = c("active", "control"),
  strata = list(site = c("north", "south"))
)
arms(trial)
```

## 2. Plan the sample size

```{r power}
# Two-arm comparison of a continuous endpoint (e.g. change in gait speed).
sampleSizeContinuous(delta = 0.12, sd = 0.20, power = 0.8)
```

## 3. Randomize with allocation concealment

Stratified permuted-block randomization is reproducible from a seed. Future
treatment assignments stay masked in the sealed sequence.

```{r randomize}
participants <- data.frame(
  id = sprintf("P%03d", 1:8),
  site = rep(c("north", "south"), each = 4)
)
sequence <- randomize(
  trial,
  method = "stratified_block",
  participants = participants,
  seed = 2026
)
assignments(sequence)
```

Allocations are revealed one at a time, through an auditable request:

```{r allocate}
allocation <- nextAllocation(sequence, requester = "site")
allocation$arm
sequence <- allocation$sequence
```

## 4. CONSORT flow

CONSORT counts and the flow diagram are inspectable without a rendering
dependency (`render = FALSE`).

```{r consort}
flow <- data.frame(
  participant_id = c("P001", "P002"),
  eligible = c(TRUE, FALSE),
  randomized = c(TRUE, FALSE),
  arm = c("active", NA),
  received_allocated = c(TRUE, NA),
  followup_complete = c(TRUE, NA),
  analysed = c(TRUE, NA),
  pre_exclusion_reason = c(NA, "ineligible"),
  not_received_reason = NA_character_,
  followup_reason = NA_character_,
  analysis_exclusion_reason = NA_character_
)
diagram <- consortDiagram(trial, flow, render = FALSE)
consortCounts(diagram)
```

## 5. Analysis populations

Intention-to-treat and per-protocol membership consume explicit finalized
assignments.

```{r populations}
participant_state <- data.frame(
  id = c("P001", "P002", "screen-01"),
  arm = c("active", "control", NA),
  randomized = c(TRUE, TRUE, FALSE),
  adherent = c(TRUE, FALSE, NA)
)
itt <- intentionToTreat(trial, participant_state)
pp  <- perProtocol(trial, participant_state, adherent_col = "adherent")
analysisExclusions(pp)
```

## 6. CDISC-shaped export

Exports keep the source-to-`USUBJID` map explicit and are structurally
validated.

```{r cdisc}
sdtm <- toSDTM(trial, participant_state)
validateCDISC(sdtm)
sdtm$metadata$id_map
```

## Where to go next

`?PhysioTrial` lists every entry point: blinding and unblinding audits
(`blindingManager()`, `unblind()`), SPIRIT schedules (`spiritSchedule()`),
adverse-event summaries (`aeSummary()`), endpoint wrappers (`endpointANCOVA()`,
`endpointMMRM()`), and the ADaM export (`toADaM()`).
